---
title: "LSCOG 2050 LRTP - Base Year SE Data Development"
author:
  - name: Pukar Bhandari
    email: pbhandari@metroanalytics.com
    affiliation:
      - name: Metro Analytics
        url: "https://www.metroanalytics.com/"
date: "2025-07-05"
engine: knitr
execute: 
  warning: false
  message: false
categories: [r, python, gis, data science, spatial analysis, transportation]
---

```{r reticulate}
#| include: false

# Setup reticulate for Python code execution
# install.packages("reticulate", dependencies = TRUE)
library(reticulate)

reticulate::use_condaenv("metro", required = TRUE)

# # Install python packages
# reticulate::py_install(
#   packages = c("numpy", "pandas", "geopandas", "shapely", "folium", "requests", "pygris"),
#   envname = "metro",
#   method = "conda",
#   pip = TRUE
# )
```

# Set up the environment

This section establishes the computational environment for processing socioeconomic data inputs for the Lower Savannah Council of Governments regional travel demand model using both R and Python platforms.

## Install and load packages

The package installation process incorporates essential libraries for comprehensive geospatial data analysis.

The R environment utilizes the `pacman` package manager to streamline the installation of multiple packages simultaneously, including `tidyverse` for data manipulation, `sf` for spatial data handling, `tidycensus` for Census Bureau data access, and `lehdr` for Longitudinal Employer-Household Dynamics data retrieval.

The Python environment focuses on core data science libraries including `pandas` for data manipulation, `geopandas` for spatial analysis, and `pygris` for Census data queries. These packages form the analytical backbone for processing demographic, employment, and geographic data required for travel demand modeling.

::: panel-tabset

#### R

```{r}
# Install and load the pacman package
if (!require("pacman")) {
  install.packages("pacman")
}
library("pacman") # Package Management Tool CRAN v0.5.1

# Install and load multiple desired packages at once
pacman::p_load(
  tidyverse, # Easily Install and Load the 'Tidyverse'
  sf, # Simple Features for R
  tidycensus, # Load US Census Boundary and Attribute Data
  lehdr, # Grab Longitudinal Employer-Household Dynamics (LEHD)
  arcgis, # ArcGIS Location Services Meta-Package
  mapview, # Interactive Viewing of Spatial Data
  RColorBrewer, # Color Palettes
  janitor # Simple Tools for Examining and Cleaning Dirty Data
)
```

#### Python

```{python}
# Install required packages if not available
# pip install numpy pandas geopandas matplotlib seaborn folium pathlib zipfile requests urllib warnings pygris

# General
import os
from pathlib import Path
import zipfile
import requests
import urllib.parse
import warnings
warnings.filterwarnings('ignore')

# Data and Visualization
import numpy as np
import pandas as pd
import geopandas as gpd
import folium
from shapely.geometry import Point

# Census data query
from pygris import blocks, block_groups
from pygris.helpers import validate_state, validate_county
from pygris.data import get_census, get_lodes
```

:::

## Set global options and parameters

Configuration settings optimize performance and establish spatial consistency. The `tigris` cache prevents redundant TIGER/Line shapefile downloads. The South Carolina State Plane coordinate system (`EPSG:3361`) serves as the standard projection for accurate GIS operations

::: panel-tabset

#### R

```{r}
# Set options
options(tigris_use_cache = TRUE) # cache tiger/line shapefile for future use

# set project CRS
project_crs <- "EPSG:3361"
```

#### Python

```{python}
# Set project CRS
project_crs = "EPSG:3361"
```

:::

## Set census API key

API authentication enables access to detailed demographic and economic datasets from the Census Bureau. The key configuration supports both R and Python environments for automated data retrieval workflows.

> 💡 **Need a Census API key?** Get one for free at [census.gov/developers](https://www.census.gov/developers/)

::: panel-tabset

#### R

```{r}
#| eval: false

# Set your API key into environment
tidycensus::census_api_key("your_api_key_here", install = TRUE)
```

#### Python

```{python}
#| eval: false

# Set your API key into environment
os.environ['CENSUS_API_KEY'] = 'your_api_key_here'
```

:::

## Project folder

The centralized directory structure organizes input data, processing files, and model outputs. The standardized `root` folder path ensures consistent file management across computing environments and team members.

::: panel-tabset

#### R

```{r}
# Set your main data folder
root <- "M:/MA_Project/SC_LSCOG LRTP"
```

#### Python

```{python}
# Set your main data folder
root = "M:/MA_Project/SC_LSCOG LRTP"
```

:::

# Define study area

This section defines the geographic extent of the Lower Savannah Council of Governments region and loads the Traffic Analysis Zone (TAZ) geometry for spatial analysis.

## Define state and counties

The study area encompasses six counties within South Carolina: Aiken, Allendale, Bamberg, Barnwell, Calhoun, and Orangeburg. These counties constitute the LSCOG planning region for travel demand modeling purposes.

::: panel-tabset

#### R

```{r}
# Define state abbreviation and county names
state_abb <- "SC"
county_names <- c(
  "Aiken",
  "Allendale",
  "Bamberg",
  "Barnwell",
  "Calhoun",
  "Orangeburg"
)
```

#### Python

```{python}
# Define state abbreviation and county names
state_abb = "SC"
county_names = [
    "Aiken",
    "Allendale",
    "Bamberg",
    "Barnwell",
    "Calhoun",
    "Orangeburg"
]
```

:::

FIPS code conversion translates state abbreviations and county names into standardized Federal Information Processing Standard codes. These codes enable consistent data retrieval across census datasets and ensure proper geographic matching with demographic and economic data sources.

::: panel-tabset

#### R

```{r}
# Converting state abbreviation code to FIPS code
state_fips <- tidycensus:::validate_state(state = state_abb)
county_fips <- vapply(
  county_names,
  function(x) tidycensus:::validate_county(state = state_abb, county = x),
  character(1)
)

# Converting County Names to FIPS code
fips_codes <- paste(state_fips, county_fips, sep = "")
fips_codes
```

#### Python

```{python}
# Converting state abbreviation code to FIPS code
state_fips = validate_state(state_abb)

# Converting County Names to FIPS code
county_fips = [
    validate_county(state_fips, county)
    for county in county_names
]

# Converting County Names to FIPS code
fips_codes = [f"{state_fips}{county}" for county in county_fips]
fips_codes
```

:::

## Load TAZ geometry

The TAZ shapefile provides the fundamental spatial framework for travel demand modeling. The geometry is loaded from the TDM exports geodatabase and filtered to include only zones within the six-county study area using FIPS code matching.

Coordinate transformation converts the TAZ geometry to the project's standard coordinate reference system (`EPSG:3361`) for accurate spatial calculations. The attribute selection retains essential fields including TAZ identifiers, area measurements, area type classifications, and county assignments.

::: panel-tabset

#### R

```{r}
# Load TAZ Shapefile
lscog_taz <- sf::read_sf(
  file.path(root, "GIS/data_temp/TDM Exports/TDM_Exports.gdb"),
  query = paste0(
    "SELECT * FROM \"SE_2019_AD_10_30_2023\" WHERE countyID IN (",
    paste0("'", fips_codes, "'", collapse = ", "),
    ")"
  )
) |>
  sf::st_transform(project_crs) |>
  dplyr::select(
    ID,
    Area,
    Acres,
    TAZ_ID = TAZ_IDs,
    AREA_TYPE,
    COUNTY,
    COUNTYID = countyID
  )

lscog_taz
```

#### Python

```{python}
# Load TAZ Shapefile
lscog_taz = gpd.read_file(
    Path(root) / "GIS/data_temp/TDM Exports/TDM_Exports.gdb",
    layer="SE_2019_AD_10_30_2023",
    where=f"countyID IN ({', '.join([f"'{fips}'" for fips in fips_codes])})"
)

lscog_taz = lscog_taz.to_crs(project_crs)

lscog_taz = lscog_taz.rename(
  columns={
    'TAZ_IDs': 'TAZ_ID',
    'countyID': 'COUNTYID'
})[['ID', 'Area', 'Acres', 'TAZ_ID', 'AREA_TYPE', 'COUNTY', 'COUNTYID', 'geometry']]

lscog_taz
```

:::

The interactive map visualization displays the TAZ structure colored by county, providing spatial context for the analysis area and enabling quality assurance of the geometric data loading process.

::: panel-tabset

#### R

```{r eval = knitr::is_html_output()}
# Create interactive map
mapview::mapview(lscog_taz, zcol = "COUNTY", lwd = 1.6, map.types = "CartoDB.Positron", col.regions = RColorBrewer::brewer.pal(6, "Dark2"))
```

#### Python

```{python eval = knitr::is_html_output()}
# Create interactive map
lscog_taz.explore(column="COUNTY", categorical=True, legend=True, tiles="CartoDB positron", zoom_start=8)
```

:::

# Fetch raw data

This section retrieves demographic, economic, and employment data from multiple Census Bureau sources at the appropriate geographic scales for travel demand modeling.

## 2020 Decennial census

The 2020 Decennial Census provides population and housing data at the census block level, offering the finest spatial resolution for demographic analysis. Population variables include total population, group quarters population, and household population derived by subtraction. Housing variables encompass total dwelling units and household counts by size categories.

Household size distributions are consolidated into four categories: 1-person, 2-person, 3-person, and 4-or-more-person households. The 4-or-more category aggregates larger household sizes to simplify model implementation while maintaining essential demographic stratification for trip generation analysis.

::: panel-tabset

#### R

```{r}
# Define variables to download
dec_variables <- c(
    TOTPOP = "P1_001N", # Total Population
    GQPOP = "P18_001N", # Population living in Group Quarters
    DU = "H1_001N", # Dwelling Units
    HH_1 = "H9_002N", # 1-person household
    HH_2 = "H9_003N", # 2-person household
    HH_3 = "H9_004N", # 3-person household
    # HH_4 = "H9_005N", # 4-person household
    # HH_5 = "H9_006N", # 5-person household
    # HH_6 = "H9_007N", # 6-person household
    # HH_7 = "H9_008N", # 7-or-more-person household
    HH = "H9_001N" # Total Number of Households
  )

# Load Population and Household Data
lscog_dec <- tidycensus::get_decennial(
  year = 2020,
  sumfile = "dhc",
  geography = "block",
  state = state_fips,
  county = county_fips,
  output = "wide",
  cb = FALSE,
  geometry = TRUE,
  keep_geo_vars = TRUE,
  # key = Sys.getenv('CENSUS_API_KEY'),
  variables = dec_variables
) |>
  sf::st_transform(project_crs) |>
  dplyr::mutate(
    HHPOP = TOTPOP - GQPOP,
    HH_4 = HH - (HH_1 + HH_2 + HH_3)
  ) |>
  dplyr::select(GEOID, TOTPOP, GQPOP, HHPOP, HH, HH_1, HH_2, HH_3, HH_4, DU)

lscog_dec
```

#### Python

```{python}
# Define variables to download
dec_variables = {
    'P1_001N': 'TOTPOP',     # Total Population
    'P18_001N': 'GQPOP',      # Population living in Group Quarters
    'H1_001N': 'DU',          # Dwelling Units
    'H9_002N': 'HH_1',        # 1-person household
    'H9_003N': 'HH_2',        # 2-person household
    'H9_004N': 'HH_3',        # 3-person household
    # 'H9_005N': 'HH_4',        # 4-person household
    # 'H9_006N': 'HH_5',        # 5-person household
    # 'H9_007N': 'HH_6',        # 6-person household
    # 'H9_008N': 'HH_7',        # 7-or-more-person household
    'H9_001N': 'HH'           # Total Number of Households
}

# get census block geometries
lscog_cb = blocks(
    state=state_fips,
    county=county_fips,
    year=2020,
    cache=True
)

# Download decennial census data at block level
lscog_dec = get_census(
    dataset="dec/dhc",
    year=2020,
    variables=list(dec_variables.keys()),
    params={
        "for": f"block:*",
        # "key": f"{os.getenv('CENSUS_API_KEY')}",
        "in": f"state:{state_fips} county:{','.join(county_fips)}"
    },
    return_geoid=True,
    guess_dtypes=True,
)

# join data to geometry
lscog_dec = lscog_cb[['GEOID20', 'geometry']].merge(lscog_dec, left_on = "GEOID20", right_on = "GEOID")

# Rename columns
lscog_dec = lscog_dec.rename(columns=dec_variables)

# Transform CRS
lscog_dec = lscog_dec.to_crs(project_crs)

# Calculate derived variables
lscog_dec['HHPOP'] = lscog_dec['TOTPOP'] - lscog_dec['GQPOP']
lscog_dec['HH_4'] = lscog_dec['HH'] - (
    lscog_dec['HH_1'] + lscog_dec['HH_2'] + lscog_dec['HH_3']
)

# Select final columns
lscog_dec = lscog_dec[['GEOID', 'TOTPOP', 'GQPOP', 'HHPOP',
                      'HH', 'HH_1', 'HH_2', 'HH_3', 'HH_4', 'DU', 'geometry']]

lscog_dec
```
:::

## 2020 ACS estimates

The American Community Survey 5-year estimates provide household income data at the block group level. Income categories are aggregated into three broad ranges: under \$15,000, \$15,000-\$49,999, and \$50,000 and above. This stratification aligns with travel behavior research indicating distinct mobility patterns across income levels.

The block group geography represents the finest spatial resolution available for ACS income data, providing sufficient detail for socioeconomic modeling while maintaining statistical reliability through the 5-year aggregation period.

::: panel-tabset

#### R

```{r}
# Define variables to download
acs_variables <- c(
    INC_CAT_02 = "B19001_002", # Less than $10,000
    INC_CAT_03 = "B19001_003", # $10,000 to $14,999
    INC_CAT_04 = "B19001_004", # $15,000 to $19,999
    INC_CAT_05 = "B19001_005", # $20,000 to $24,999
    INC_CAT_06 = "B19001_006", # $25,000 to $29,999
    INC_CAT_07 = "B19001_007", # $30,000 to $34,999
    INC_CAT_08 = "B19001_008", # $35,000 to $39,999
    INC_CAT_09 = "B19001_009", # $40,000 to $44,999
    INC_CAT_10 = "B19001_010", # $45,000 to $49,999
    # INC_CAT_11 = "B19001_011", # $50,000 to $59,999
    # INC_CAT_12 = "B19001_012", # $60,000 to $74,999
    # INC_CAT_13 = "B19001_013", # $75,000 to $99,999
    # INC_CAT_14 = "B19001_014", # $100,000 to $124,999
    # INC_CAT_15 = "B19001_015", # $125,000 to $149,999
    # INC_CAT_16 = "B19001_016", # $150,000 to $199,999
    # INC_CAT_17 = "B19001_017", # $200,000 or more
    INC_CAT_01 = "B19001_001" # Total
  )

# Load Household Income Data
lscog_acs <- tidycensus::get_acs(
  year = 2020,
  survey = "acs5",
  geography = "block group",
  state = state_fips,
  county = county_fips,
  output = "wide",
  cb = FALSE,
  geometry = TRUE,
  # key = Sys.getenv('CENSUS_API_KEY'),
  variables = acs_variables
) |>
  sf::st_transform(project_crs) |>
  dplyr::mutate(
    INC_14999 = INC_CAT_02E + INC_CAT_03E,
    INC_49999 = INC_CAT_04E +
      INC_CAT_05E +
      INC_CAT_06E +
      INC_CAT_07E +
      INC_CAT_08E +
      INC_CAT_09E +
      INC_CAT_10E,
    INC_50000 = INC_CAT_01E - (INC_14999 + INC_49999)
  ) |>
  dplyr::select(GEOID, INC_TOTAL = INC_CAT_01E, INC_14999, INC_49999, INC_50000)

lscog_acs
```

#### Python

```{python}
# Define variables to download
acs_variables = {
    'B19001_002E': 'INC_CAT_02',  # Less than $10,000
    'B19001_003E': 'INC_CAT_03',  # $10,000 to $14,999
    'B19001_004E': 'INC_CAT_04',  # $15,000 to $19,999
    'B19001_005E': 'INC_CAT_05',  # $20,000 to $24,999
    'B19001_006E': 'INC_CAT_06',  # $25,000 to $29,999
    'B19001_007E': 'INC_CAT_07',  # $30,000 to $34,999
    'B19001_008E': 'INC_CAT_08',  # $35,000 to $39,999
    'B19001_009E': 'INC_CAT_09',  # $40,000 to $44,999
    'B19001_010E': 'INC_CAT_10',  # $45,000 to $49,999
    # 'B19001_011E': 'INC_CAT_11',  # $50,000 to $59,999
    # 'B19001_012E': 'INC_CAT_12',  # $60,000 to $74,999
    # 'B19001_013E': 'INC_CAT_13',  # $75,000 to $99,999
    # 'B19001_014E': 'INC_CAT_14',  # $100,000 to $124,999
    # 'B19001_015E': 'INC_CAT_15',  # $125,000 to $149,999
    # 'B19001_016E': 'INC_CAT_16',  # $150,000 to $199,999
    # 'B19001_017E': 'INC_CAT_17',  # $200,000 or more
    'B19001_001E': 'INC_CAT_01'   # Total
}

# get blockgroup geometries
lscog_bg = block_groups(
    state=state_fips,
    county=county_fips,
    year=2020,
    cache=True
)

# Download household income data at block group level
lscog_acs = get_census(
    dataset="acs/acs5",
    year=2020,
    variables=list(acs_variables.keys()),
    params={
        "for": f"block group:*",
        # "key": f"{os.getenv('CENSUS_API_KEY')}",
        "in": f"state:{state_fips} county:{','.join(county_fips)}"
    },
    return_geoid=True,
    guess_dtypes=True
)

# join data to geometry
lscog_acs = lscog_bg[['GEOID', 'geometry']].merge(lscog_acs, on = "GEOID")

# Rename columns
lscog_acs = lscog_acs.rename(columns=acs_variables)

# Transform CRS
lscog_acs = lscog_acs.to_crs(project_crs)

# Calculate derived variables
lscog_acs['INC_14999'] = lscog_acs['INC_CAT_02'] + lscog_acs['INC_CAT_03']
lscog_acs['INC_49999'] = (
    lscog_acs['INC_CAT_04'] +
    lscog_acs['INC_CAT_05'] +
    lscog_acs['INC_CAT_06'] +
    lscog_acs['INC_CAT_07'] +
    lscog_acs['INC_CAT_08'] +
    lscog_acs['INC_CAT_09'] +
    lscog_acs['INC_CAT_10']
)
lscog_acs['INC_50000'] = lscog_acs['INC_CAT_01'] - (
    lscog_acs['INC_14999'] + lscog_acs['INC_49999']
)

# Select final columns
lscog_acs = lscog_acs.rename(columns={'INC_CAT_01': 'INC_TOTAL'})
lscog_acs = lscog_acs[['GEOID', 'INC_TOTAL', 'INC_14999', 'INC_49999', 'INC_50000', 'geometry'
]]

lscog_acs
```
:::

## 2019 LEHD data

The Longitudinal Employer-Household Dynamics Workplace Area Characteristics data provides employment counts by industry sector at the census block level. Employment categories follow the North American Industry Classification System and are aggregated into transportation-relevant sectors including retail, services, manufacturing, and public administration.

The 2019 reference year represents pre-pandemic employment patterns, providing a stable baseline for long-term transportation planning. Employment data at the block level enables precise spatial allocation of work destinations within the travel demand model framework.

::: panel-tabset
#### R

```{r}

# Download LEHD WAC data at block level
lscog_emp <- lehdr::grab_lodes(
  version = "LODES8",
  state = tolower(state_abb),
  lodes_type = "wac",
  segment = "S000",
  job_type = "JT00",
  year = 2019,
  state_part = "",
  agg_geo = "block",
  use_cache = TRUE
) |>
  dplyr::filter(grepl(
    paste("^(", paste(fips_codes, collapse = "|"), ")", sep = ""),
    w_geocode
  )) |>
  # check the documentation at: https://lehd.ces.census.gov/data/lodes/LODES8/LODESTechDoc8.0.pdf
  dplyr::mutate(
    GEOID = as.character(w_geocode),
    TOTAL_EMP = C000, # Total Employment
    AGR_FOR_FI = CNS01, # Agricultural, forestry, and fishing employment
    MINING = CNS02, # Mining employment
    CONSTRUCTI = CNS04, # Construction employment
    MANUFACTUR = CNS05, # Manufacturing employment
    TRANSP_COM = CNS08 + CNS09, # Transportation, communication employment
    WHOLESALE = CNS06, # Wholesale employment
    RETAIL = CNS07, # Retail employment
    FIRE = CNS10 + CNS11, # Finance / Insurance / Real Estate employment
    SERVICES = CNS03 +
      CNS12 +
      CNS13 +
      CNS14 + # Service employment
      CNS15 +
      CNS16 +
      CNS17 +
      CNS18 +
      CNS19,
    PUBLIC_ADM = CNS20 # Public Administration employment
  ) |>
  dplyr::select(
    GEOID,
    TOTAL_EMP,
    AGR_FOR_FI,
    MINING,
    CONSTRUCTI,
    MANUFACTUR,
    TRANSP_COM,
    WHOLESALE,
    RETAIL,
    FIRE,
    SERVICES,
    PUBLIC_ADM
  )

lscog_emp
```

#### Python

```{python}
# Download LEHD WAC data at block level
lscog_emp = get_lodes(
    state=state_abb,
    year=2019,
    version="LODES8",
    lodes_type="wac",
    part="main",
    segment="S000",
    job_type="JT00",
    agg_level="block",
    cache=True,
    return_geometry=False
)

# Filter for specific FIPS codes
lscog_emp = lscog_emp[lscog_emp['w_geocode'].str.match(f"^({'|'.join(fips_codes)})")]

# Create new columns with employment categories
# Check documentation at: https://lehd.ces.census.gov/data/lodes/LODES8/LODESTechDoc8.0.pdf
lscog_emp = lscog_emp.assign(
    GEOID=lscog_emp['w_geocode'].astype(str),
    TOTAL_EMP=lscog_emp['C000'],  # Total Employment
    AGR_FOR_FI=lscog_emp['CNS01'],  # Agricultural, forestry, and fishing employment
    MINING=lscog_emp['CNS02'],  # Mining employment
    CONSTRUCTI=lscog_emp['CNS04'],  # Construction employment
    MANUFACTUR=lscog_emp['CNS05'],  # Manufacturing employment
    TRANSP_COM=lscog_emp['CNS08'] + lscog_emp['CNS09'],  # Transportation, communication employment
    WHOLESALE=lscog_emp['CNS06'],  # Wholesale employment
    RETAIL=lscog_emp['CNS07'],  # Retail employment
    FIRE=lscog_emp['CNS10'] + lscog_emp['CNS11'],  # Finance / Insurance / Real Estate employment
    SERVICES=(lscog_emp['CNS03'] +
              lscog_emp['CNS12'] +
              lscog_emp['CNS13'] +
              lscog_emp['CNS14'] +
              lscog_emp['CNS15'] +
              lscog_emp['CNS16'] +
              lscog_emp['CNS17'] +
              lscog_emp['CNS18'] +
              lscog_emp['CNS19']),  # Service employment
    PUBLIC_ADM=lscog_emp['CNS20']  # Public Administration employment
)

# Select only the desired columns
lscog_emp = lscog_emp[['GEOID', 'TOTAL_EMP', 'AGR_FOR_FI', 'MINING', 'CONSTRUCTI',
                       'MANUFACTUR', 'TRANSP_COM', 'WHOLESALE', 'RETAIL', 'FIRE',
                       'SERVICES', 'PUBLIC_ADM']]

# Display structure/info about the dataframe
lscog_emp
```
:::

## 2020 NCES school and college enrollment data

The National Center for Education Statistics provides comprehensive educational institution data including enrollment and staffing information for transportation planning analysis.

### Public schools

Public school data is retrieved from the NCES ArcGIS REST service for the 2019-2020 academic year. The dataset includes total student enrollment and full-time equivalent teacher counts for each institution within the six-county region. Public schools represent major trip generation sources for both student and employee travel, requiring precise spatial location data for accurate modeling.

::: panel-tabset

#### R

```{r}

# Public School Location data 2019-2020
lscog_pub_sch_enroll <- arcgislayers::arc_read(
  url = "https://nces.ed.gov/opengis/rest/services/K12_School_Locations/EDGE_ADMINDATA_PUBLICSCH_1920/MapServer/0",
  where = paste0(
    "LSTATE = '",
    state_abb,
    "' AND NMCNTY IN (",
    paste0("'", paste0(county_names, " County"), "'", collapse = ", "),
    ")"
  ),
  alias = "label",
  crs = project_crs
) |>
  dplyr::select(
    INSTITUTION_ID = NCESSCH,
    NAME = SCH_NAME,
    STATE = LSTATE,
    STUDENT_COUNT_PUB = TOTAL,
    TEACHER_COUNT_PUB = FTE
  )

lscog_pub_sch_enroll
```

#### Python

```{python}
#| code-fold: true

# Create function to read ArcGIS FeatureLayer or Table
def arc_read(url, where="1=1", outFields="*", outSR=4326, **kwargs):
    """
    Read an ArcGIS FeatureLayer or Table to a GeoDataFrame.

    Parameters:
    url (str): The ArcGIS REST service URL ending with /MapServer/0 or /FeatureServer/0
    where (str): SQL WHERE clause for filtering. Default: "1=1" (all records)
    outFields (str): Comma-separated field names or "*" for all fields. Default: "*"
    outSR (int): Output spatial reference EPSG code. Default: 4326
    **kwargs: Additional query parameters passed to the ArcGIS REST API

    Returns:
    geopandas.GeoDataFrame: Spatial data from the service
    """

    # Ensure URL ends with /query
    if not url.endswith('/query'):
        url = url.rstrip('/') + '/query'

    # Build query parameters
    params = {
        'where': where,
        'outFields': outFields,
        'returnGeometry': 'true',
        # 'geometryType': 'esriGeometryPoint',
        'outSR': outSR,
        'f': 'geojson'
    }

    # Add any additional parameters
    params.update(kwargs)

    # Make request
    response = requests.get(url, params=params)

    # Read as GeoDataFrame
    return gpd.read_file(response.text)
```

```{python}
# Public School Enrollment data 2019-2020
lscog_pub_sch_enroll = arc_read(
    url="https://nces.ed.gov/opengis/rest/services/K12_School_Locations/EDGE_ADMINDATA_PUBLICSCH_1920/MapServer/0",
    where=f"LSTATE = '{state_abb}' AND NMCNTY IN ('{"', '".join([f"{name} County" for name in county_names])}')",
    outFields='NCESSCH,SCH_NAME,LSTATE,TOTAL,FTE'
)

# Transform CRS
lscog_pub_sch_enroll = lscog_pub_sch_enroll.to_crs(project_crs)

# Select and rename columns
lscog_pub_sch_enroll = lscog_pub_sch_enroll.rename(columns={
    'NCESSCH': 'INSTITUTION_ID',
    'SCH_NAME': 'NAME',
    'LSTATE': 'STATE',
    'TOTAL': 'STUDENT_COUNT_PUB',
    'FTE': 'TEACHER_COUNT_PUB'
})

lscog_pub_sch_enroll
```
:::

### Private schools

Private school enrollment data is accessed from the NCES Private School Survey archived dataset. The data is spatially enabled using latitude and longitude coordinates and filtered to include only institutions within the study area TAZ boundaries. Private schools contribute to the regional education trip matrix and must be incorporated alongside public institutions for comprehensive coverage.

::: panel-tabset

#### R
```{r}
# Private School Enrollment data 2019-2020
lscog_pvt_sch_enroll <- vroom::vroom(
  unz(
    file.path(
      root,
      "GIS/data_external/20250315 NCES/PSS - Private/2019-20/pss1920_pu_csv.zip"
    ),
    "pss1920_pu.csv"
  ),
  col_types = vroom::cols_only(
    PPIN        = vroom::col_character(),
    PINST       = vroom::col_character(),
    PL_STABB    = vroom::col_character(),
    PCNTNM      = vroom::col_character(),
    SIZE        = vroom::col_double(),
    NUMTEACH    = vroom::col_double(),
    LATITUDE20  = vroom::col_double(),
    LONGITUDE20 = vroom::col_double()
  )
) |>
  sf::st_as_sf(coords = c("LONGITUDE20", "LATITUDE20"), crs = "EPSG:4326") |>
  sf::st_transform(project_crs) |>
  sf::st_filter(lscog_taz, .predicate = st_intersects) |>
  dplyr::select(
    INSTITUTION_ID = PPIN,
    NAME = PINST,
    STATE = PL_STABB,
    STUDENT_COUNT_PVT = SIZE,
    TEACHER_COUNT_PVT = NUMTEACH
  )

lscog_pvt_sch_enroll
```

#### Python

```{python}
# Private School Enrollment data 2019-2020
zip_path = Path(root) / "GIS/data_external/20250315 NCES/PSS - Private/2019-20/pss1920_pu_csv.zip"

with zipfile.ZipFile(zip_path, 'r') as zip_ref:
    with zip_ref.open('pss1920_pu.csv') as csv_file:
        lscog_pvt_sch_enroll = pd.read_csv(
            csv_file,
            usecols=['PPIN', 'PINST', 'PL_STABB', 'PCNTNM', 'SIZE', 'NUMTEACH', 'LATITUDE20', 'LONGITUDE20'],
            dtype={'PPIN': 'str', 'PINST': 'str', 'PL_STABB': 'str', 'PCNTNM': 'str', 'SIZE': 'float64', 'NUMTEACH': 'float64'}
        )

lscog_pvt_sch_enroll = gpd.GeoDataFrame(
    lscog_pvt_sch_enroll,
    geometry=gpd.points_from_xy(lscog_pvt_sch_enroll['LONGITUDE20'], lscog_pvt_sch_enroll['LATITUDE20']),
    crs='EPSG:4326'
).to_crs(project_crs)

lscog_pvt_sch_enroll = gpd.sjoin(lscog_pvt_sch_enroll, lscog_taz, how='inner', predicate='intersects')[
    ['PPIN', 'PINST', 'PL_STABB', 'SIZE', 'NUMTEACH', 'geometry']
].rename(columns={
    'PPIN': 'INSTITUTION_ID',
    'PINST': 'NAME',
    'PL_STABB': 'STATE',
    'SIZE': 'STUDENT_COUNT_PVT',
    'NUMTEACH': 'TEACHER_COUNT_PVT'
})

lscog_pvt_sch_enroll
```
:::

### Post-secondary institutions

Post-secondary institution locations are obtained from the NCES Postsecondary School Locations service, filtered by state and county FIPS codes. These institutions generate significant travel demand through student commuting, employee travel, and visitor trips, making them essential components of the regional transportation network analysis.

::: panel-tabset

#### R
```{r}
# Post-Secondary Location data 2019-2020
lscog_college_loc <- arcgislayers::arc_read(
  url = "https://nces.ed.gov/opengis/rest/services/Postsecondary_School_Locations/EDGE_GEOCODE_POSTSECONDARYSCH_1920/MapServer/0",
  where = paste0(
    "STATE = '",
    state_abb,
    "' AND CNTY IN (",
    paste0("'", fips_codes, "'", collapse = ", "),
    ")"
  ),
  alias = "label",
  crs = project_crs
)

lscog_college_loc
```

#### Python
```{python}
# Post-Secondary Location data 2019-2020
lscog_college_loc = arc_read(
    url="https://nces.ed.gov/opengis/rest/services/Postsecondary_School_Locations/EDGE_GEOCODE_POSTSECONDARYSCH_1920/MapServer/0",
    where=f"STATE = '{state_abb}' AND CNTY IN ('{"', '".join([f"{fip}" for fip in fips_codes])}')",
    outFields='*',
    outSR=project_crs
)

lscog_college_loc = lscog_college_loc.to_crs(project_crs)

lscog_college_loc
```
:::

# Clean data

The data cleaning process involves harmonizing multiple Census data sources to create a comprehensive socioeconomic dataset at the census block level. This requires careful interpolation and integration of American Community Survey (ACS) estimates with Decennial Census counts to maintain spatial consistency and statistical accuracy.

## Household-weighted interpolation

The interpolation process transfers ACS block group data to individual census blocks using household counts as weights. This method ensures that socioeconomic characteristics are distributed proportionally based on residential density rather than simple geometric overlay. The tidycensus package provides robust interpolation functionality that preserves the extensive nature of count variables while maintaining spatial relationships.

:::panel-tabset

#### R
```{r}
# Interpolate ACS data to Decennial Census blocks
lscog_acs_cb <- tidycensus::interpolate_pw(
  from = lscog_acs,
  to = lscog_dec,
  to_id = "GEOID",
  extensive = TRUE,
  weights = lscog_dec,
  crs = project_crs,
  weight_column = "HH"
)

lscog_acs_cb
```

#### Python

```{python}
#| code-fold: true

def interpolate_pw(from_gdf, to_gdf, weights_gdf, to_id=None, extensive=True, 
                   weight_column=None, weight_placement='surface', crs=None):
    """
    Population-weighted areal interpolation between geometries.
    
    Transfers numeric data from source geometries to target geometries using
    population-weighted interpolation based on point weights (e.g., census blocks).
    
    Parameters:
    -----------
    from_gdf : GeoDataFrame
        Source geometries with numeric data to interpolate
    to_gdf : GeoDataFrame  
        Target geometries to interpolate data to
    weights_gdf : GeoDataFrame
        Weight geometries (e.g., census blocks) used for interpolation.
        If polygons, will be converted to points. Can be the same as to_gdf.
    to_id : str, optional
        Column name for unique identifier in target geometries.
        If None, creates an 'id' column.
    extensive : bool, default True
        If True, return weighted sums (for counts).
        If False, return weighted means (for rates/percentages).
    weight_column : str, optional
        Column name in weights_gdf for weighting (e.g., 'POP', 'HH').
        If None, all weights are equal.
    weight_placement : str, default 'surface'
        How to convert polygons to points: 'surface' or 'centroid'
    crs : str or CRS object, optional
        Coordinate reference system to project all datasets to
        
    Returns:
    --------
    GeoDataFrame
        Target geometries with interpolated numeric values
    """
    
    # Input validation
    if not all(isinstance(gdf, gpd.GeoDataFrame) for gdf in [from_gdf, to_gdf, weights_gdf]):
        raise ValueError("All inputs must be GeoDataFrames")
    
    # Make copies to avoid modifying originals
    from_gdf = from_gdf.copy()
    to_gdf = to_gdf.copy()
    weights_gdf = weights_gdf.copy()
    
    # Set CRS if provided
    if crs:
        from_gdf = from_gdf.to_crs(crs)
        to_gdf = to_gdf.to_crs(crs)
        weights_gdf = weights_gdf.to_crs(crs)
    
    # Check CRS consistency
    if not (from_gdf.crs == to_gdf.crs == weights_gdf.crs):
        raise ValueError("All inputs must have the same CRS")
    
    # Handle to_id
    if to_id is None:
        to_id = 'id'
        to_gdf[to_id] = to_gdf.index.astype(str)
    
    # Remove conflicting columns
    if to_id in from_gdf.columns:
        from_gdf = from_gdf.drop(columns=[to_id])
    
    # Create unique from_id
    from_id = 'from_id'
    from_gdf[from_id] = from_gdf.index.astype(str)
    
    # Handle weight column
    if weight_column is None:
        weight_column = 'interpolation_weight'
        weights_gdf[weight_column] = 1.0
    else:
        # Rename to avoid conflicts
        weights_gdf['interpolation_weight'] = weights_gdf[weight_column]
        weight_column = 'interpolation_weight'
    
    # Convert weights to points if needed
    if weights_gdf.geometry.geom_type.iloc[0] in ['Polygon', 'MultiPolygon']:
        if weight_placement == 'surface':
            weights_gdf = weights_gdf.copy()
            weights_gdf.geometry = weights_gdf.geometry.representative_point()
        elif weight_placement == 'centroid':
            weights_gdf = weights_gdf.copy()
            weights_gdf.geometry = weights_gdf.geometry.centroid
        else:
            raise ValueError("weight_placement must be 'surface' or 'centroid'")
    
    # Keep only weight column and geometry
    weight_points = weights_gdf[[weight_column, 'geometry']].copy()
    
    # Calculate denominators (total weights per source geometry)
    with warnings.catch_warnings():
        warnings.filterwarnings('ignore', category=UserWarning)
        source_weights = gpd.sjoin(from_gdf, weight_points, how='left', predicate='contains')
    
    denominators = (source_weights.groupby(from_id)[weight_column]
                   .sum()
                   .reset_index()
                   .rename(columns={weight_column: 'weight_total'}))
    
    # Calculate intersections between from and to
    with warnings.catch_warnings():
        warnings.filterwarnings('ignore', category=UserWarning)
        intersections = gpd.overlay(from_gdf, to_gdf, how='intersection')
    
    # Filter to keep only polygon intersections
    intersections = intersections[intersections.geometry.geom_type.isin(['Polygon', 'MultiPolygon', 'GeometryCollection'])]
    
    if len(intersections) == 0:
        raise ValueError("No valid polygon intersections found between source and target geometries")
    
    # Add intersection ID
    intersections['intersection_id'] = range(len(intersections))
    
    # Spatial join intersections with weight points to get weights within each intersection
    with warnings.catch_warnings():
        warnings.filterwarnings('ignore', category=UserWarning)
        intersection_weights = gpd.sjoin(intersections, weight_points, how='left', predicate='contains')
    
    # Calculate intersection values (sum of weights per intersection)
    intersection_values = (intersection_weights.groupby('intersection_id')[weight_column]
                         .sum()
                         .reset_index()
                         .rename(columns={weight_column: 'intersection_value'}))
    
    # Merge back to intersections and keep only unique intersections
    intersections = intersections.merge(intersection_values, on='intersection_id', how='left')
    intersections['intersection_value'] = intersections['intersection_value'].fillna(0)
    
    # Remove duplicates created by the spatial join
    intersections = intersections.drop_duplicates(subset='intersection_id')
    
    # Merge with denominators to calculate weight coefficients
    intersections = intersections.merge(denominators, on=from_id, how='left')
    intersections['weight_total'] = intersections['weight_total'].fillna(1)
    
    # Calculate weight coefficients (intersection weight / total weight in source)
    intersections.loc[intersections['weight_total'] > 0, 'weight_coef'] = (
        intersections['intersection_value'] / intersections['weight_total']
    )
    intersections['weight_coef'] = intersections['weight_coef'].fillna(0)
    
    # Get numeric columns from source data
    numeric_cols = from_gdf.select_dtypes(include=[np.number]).columns
    # Remove ID columns
    numeric_cols = [col for col in numeric_cols if col not in [from_id]]
    
    # Prepare intersection data for interpolation
    intersection_data = intersections[[from_id, to_id, 'weight_coef'] + numeric_cols].copy()
    
    if extensive:
        # For extensive variables: multiply by weight coefficient, then sum by target
        for col in numeric_cols:
            intersection_data[col] = intersection_data[col] * intersection_data['weight_coef']
        
        interpolated = (intersection_data.groupby(to_id)[numeric_cols]
                       .sum()
                       .reset_index())
    else:
        # For intensive variables: weighted average
        interpolated_data = []
        for target_id in intersection_data[to_id].unique():
            target_data = intersection_data[intersection_data[to_id] == target_id]
            if len(target_data) > 0 and target_data['weight_coef'].sum() > 0:
                weighted_vals = {}
                for col in numeric_cols:
                    weighted_vals[col] = (target_data[col] * target_data['weight_coef']).sum() / target_data['weight_coef'].sum()
                weighted_vals[to_id] = target_id
                interpolated_data.append(weighted_vals)
        
        interpolated = pd.DataFrame(interpolated_data)
    
    # Merge with target geometries
    result = to_gdf[[to_id, 'geometry']].merge(interpolated, on=to_id, how='left')
    
    # Fill NaN values with 0 for missing interpolations
    for col in numeric_cols:
        if col in result.columns:
            result[col] = result[col].fillna(0)
    
    return result
```

```{python}

# Interpolate ACS data to Decennial Census blocks
lscog_acs_cb = interpolate_pw(
    from_gdf=lscog_acs,
    to_gdf=lscog_dec,
    weights_gdf=lscog_dec,
    to_id='GEOID',
    extensive=True,
    weight_column='HH',
    crs=project_crs
)

lscog_acs_cb.head()
```

:::

The comparative visualization reveals the increased spatial resolution achieved through interpolation. Block-level data provides more granular detail for transportation modeling applications, enabling better representation of local variations in income distribution across the study area.

:::panel-tabset

#### R
```{r}
#| eval: false

# Compare before and after interpolation
mapview::mapview(lscog_acs_cb, zcol = "INC_49999", color = NA) |
  mapview::mapview(lscog_acs, zcol = "INC_49999", color = NA)
```

#### Python
```{python}
#| eval: false

# Compare before and after interpolation
lscog_acs_cb.explore(column="INC_49999", color="blue", legend=True, tiles="CartoDB positron") |\
    lscog_acs.explore(column="INC_49999", color="red", legend=True, tiles="CartoDB positron")
```

:::

## Combine population and households

The integration step merges the interpolated ACS socioeconomic data with the Decennial Census population and household counts. This join operation creates a unified dataset containing both demographic totals and detailed characteristics at the census block level. The left join ensures that all census blocks retain their geographic boundaries while incorporating available socioeconomic attributes.

:::panel-tabset

#### R
```{r}
## Combine ACS Data to Decennial data
lscog_pop_hh <- lscog_dec |>
  dplyr::left_join(
    sf::st_drop_geometry(lscog_acs_cb),
    by = dplyr::join_by(GEOID)
  )

lscog_pop_hh
```

#### Python
```{python}

## Combine ACS Data to Decennial data
lscog_pop_hh = lscog_dec.merge(
    lscog_acs_cb.drop(columns=['geometry']),
    on='GEOID',
    how='left'
)

lscog_pop_hh
```
:::

Income category adjustments reconcile the interpolated ACS estimates with actual household counts from the Decennial Census. The proportional allocation method redistributes income categories based on the ratio of interpolated totals to observed household counts, maintaining consistency between data sources. The three-tier income classification (under \$15,000, \$15,000-\$49,999, and \$50,000 and above) provides sufficient granularity for travel demand modeling while ensuring statistical reliability.

:::panel-tabset

#### R
```{r}

## Combine adjusted HH income level to Decennial census instead of ACS
lscog_pop_hh <- lscog_pop_hh |>
  dplyr::mutate(
    INC_49999 = tidyr::replace_na(round(INC_49999 / INC_TOTAL * HH, 0), 0),
    INC_50000 = tidyr::replace_na(round(INC_50000 / INC_TOTAL * HH, 0), 0),
    INC_14999 = HH - (INC_49999 + INC_50000)
  ) |>
  dplyr::select(-INC_TOTAL)

lscog_pop_hh
```

#### Python

```{python}

## Combine adjusted HH income level to Decennial census instead of ACS
lscog_pop_hh["INC_49999"] = ((lscog_pop_hh["INC_49999"] / lscog_pop_hh["INC_TOTAL"]) * lscog_pop_hh["HH"]).round().fillna(0)
lscog_pop_hh["INC_50000"] = ((lscog_pop_hh["INC_50000"] / lscog_pop_hh["INC_TOTAL"]) * lscog_pop_hh["HH"]).round().fillna(0)
lscog_pop_hh["INC_14999"] = lscog_pop_hh["HH"] - (lscog_pop_hh["INC_49999"] + lscog_pop_hh["INC_50000"])

lscog_pop_hh = lscog_pop_hh.drop(columns="INC_TOTAL")
lscog_pop_hh
```

:::

## Employment data

The employment integration incorporates LEHD (Longitudinal Employer-Household Dynamics) workplace area characteristics into the combined dataset. This addition provides employment counts by census block, enabling the development of trip attraction models and work-based travel pattern analysis. The merge operation maintains the geographic integrity of census blocks while adding employment variables essential for comprehensive transportation planning.

:::panel-tabset

#### R

```{r}

# Join LEHD Data to the Decennial data
lscog_pop_hh_emp <- lscog_pop_hh |>
  dplyr::left_join(lscog_emp, by = dplyr::join_by(GEOID))

lscog_pop_hh_emp
```

#### Python

```{python}

# Join LEHD Data to the Decennial data
lscog_pop_hh_emp = lscog_pop_hh.merge(
    lscog_emp,
    on='GEOID',
    how='left'
)

lscog_pop_hh_emp
```

:::

# Export raw data

The data export process creates standardized datasets for travel demand model development, maintaining both tabular and spatial formats to support various modeling applications. All exports follow consistent file naming conventions and directory structures to facilitate model integration and data management workflows.

## TAZ data

The Traffic Analysis Zone (TAZ) boundary export provides the fundamental geographic framework for the regional travel demand model. The blank TAZ file serves as a template for subsequent socioeconomic data allocation, containing only zone identification fields and geometric boundaries without attribute data.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_taz |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_taz_blank.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_taz.to_csv(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_taz_blank.csv",
    index=False
)
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R
```{r}
#| eval: false

# Export as GDB
lscog_taz |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_taz_blank",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_taz.to_file(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb",
    layer='lscog_taz_blank',
    driver='FileGDB'
)
```

:::

## Census blocks to TAZ conversion

The block-to-TAZ conversion table establishes the critical linkage between fine-scale Census geography and the modeling zone system. This crosswalk file enables the aggregation of block-level socioeconomic data to TAZ boundaries while maintaining traceability to source geographies.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_cb |>
  sf::st_join(lscog_taz) |>
  dplyr::select(GEOID20, ID, TAZ_ID) |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_block_taz.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_cb.merge(lscog_taz[['ID', 'TAZ_ID']], left_on='GEOID20', right_on='ID')[['GEOID20', 'ID', 'TAZ_ID']].to_csv(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_block_taz.csv",
    index=False
)
```

:::

## Decennial census data at census block level

The decennial census block export captures the foundational demographic counts used throughout the modeling process. This dataset provides the most reliable population and household totals at the finest geographic resolution, serving as the base for all subsequent data integration and validation steps.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_dec |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_dec_block.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_dec.to_csv(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_dec_block.csv",
    index=False
)
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as GDB
lscog_dec |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_dec_block",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_dec.to_file(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb",
    layer='lscog_dec_block',
    driver='FileGDB'
)
```

:::

## ACS estimates at census block level

The interpolated ACS data export delivers income distribution estimates at the census block level, providing the socioeconomic stratification necessary for trip generation modeling. This processed dataset represents the final product of the household-weighted interpolation methodology, ready for direct integration into the travel demand model framework.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_acs_cb |>
  dplyr::select(GEOID, INC_14999, INC_49999, INC_50000) |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_acs_block.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_acs_cb[['GEOID', 'INC_14999', 'INC_49999', 'INC_50000']].to_csv(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_acs_block.csv",
    index=False
)
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as GDB
lscog_acs_cb |>
  dplyr::select(GEOID, INC_14999, INC_49999, INC_50000) |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_acs_block",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_acs_cb[['GEOID', 'INC_14999', 'INC_49999', 'INC_50000']].to_file(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb",
    layer='lscog_acs_block',
    driver='FileGDB'
)
```

:::

## LEHD data at census block level

The employment data export provides comprehensive workplace characteristics by industry sector at the census block level. This dataset captures the spatial distribution of employment opportunities across the study region, supporting both trip attraction modeling and economic impact analysis.

**Export as CSV flat file**

:::panel-tabset

#### R
```{r}
#| eval: false

# Export as CSV
lscog_pop_hh_emp |>
  dplyr::select(GEOID, TOTAL_EMP:PUBLIC_ADM) |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_emp_block.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_pop_hh_emp[['GEOID', 'TOTAL_EMP', 'AGR_FOR_FI', 'MINING', 'CONSTRUCTI',
                   'MANUFACTUR', 'TRANSP_COM', 'WHOLESALE', 'RETAIL', 'FIRE',
                   'SERVICES', 'PUBLIC_ADM']].to_csv(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_emp_block.csv",
    index=False
)
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as GDB
lscog_pop_hh_emp |>
  dplyr::select(GEOID, TOTAL_EMP:PUBLIC_ADM) |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_emp_block",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_pop_hh_emp[['GEOID', 'TOTAL_EMP', 'AGR_FOR_FI', 'MINING', 'CONSTRUCTI',
                   'MANUFACTUR', 'TRANSP_COM', 'WHOLESALE', 'RETAIL', 'FIRE',
                   'SERVICES', 'PUBLIC_ADM']].to_file(
        Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb",
        layer='lscog_emp_block',
        driver='FileGDB'
    )
```

:::

## Public schools to TAZ

The public school location export integrates educational facility data with the TAZ system, providing essential inputs for school-related trip modeling. Student and teacher counts by facility support the development of specialized trip generation rates for educational purposes.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_pub_sch_enroll |>
  sf::st_join(lscog_taz |> dplyr::select(ID, TAZ_ID)) |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_pubsch_loc.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_pub_sch_enroll.sjoin(
    lscog_taz[['ID', 'TAZ_ID']],
    how='left'
)[['INSTITUTION_ID', 'NAME', 'STATE', 'STUDENT_COUNT_PUB', 'TEACHER_COUNT_PUB', 'geometry']] \
    .to_csv(
        Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_pubsch_loc.csv",
        index=False
    )
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as GDB
lscog_pub_sch_enroll |>
  sf::st_join(lscog_taz |> dplyr::select(ID, TAZ_ID)) |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_pubsch_loc",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_pub_sch_enroll.sjoin(
    lscog_taz[['ID', 'TAZ_ID']],
    how='left'
).to_file(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb",
    layer='lscog_pubsch_loc',
    driver='FileGDB'
)
```

:::

## Private schools to TAZ

The private school dataset complements the public education data by capturing enrollment patterns in private educational institutions. This comprehensive coverage of educational facilities ensures that all school-related travel demand is properly represented in the regional model.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_pvt_sch_enroll |>
  sf::st_join(lscog_taz |> dplyr::select(ID, TAZ_ID)) |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_raw/lscog_pvtsch_loc.csv"
    ),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_pvt_sch_enroll.sjoin(
    lscog_taz[['ID', 'TAZ_ID']],
    how='left'
)[['INSTITUTION_ID', 'NAME', 'STATE', 'STUDENT_COUNT_PVT', 'TEACHER_COUNT_PVT', 'geometry']] \
    .to_csv(
        Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_pvtsch_loc.csv",
        index=False
    )
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as GDB
lscog_pvt_sch_enroll |>
  sf::st_join(lscog_taz |> dplyr::select(ID, TAZ_ID)) |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_pvtsch_loc",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_pvt_sch_enroll.sjoin(
    lscog_taz[['ID', 'TAZ_ID']],
    how='left'
).to_file(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb",
    layer='lscog_pvtsch_loc',
    driver='FileGDB'
)
```

:::

# Aggregate data to TAZ

The aggregation process transforms fine-scale census block data into TAZ-level inputs suitable for travel demand modeling. This spatial aggregation preserves the total counts while organizing data according to the modeling zone structure required for trip generation and distribution analysis.

## Population, households, and employment

The spatial join operation aggregates all demographic, housing, and employment variables from census blocks to their corresponding TAZs using centroid-based assignment. This process ensures that each block's socioeconomic characteristics are properly allocated to the appropriate modeling zone while maintaining data integrity through comprehensive summation of all relevant variables.

:::panel-tabset

#### R

```{r}
# Aggregate population, households, and employment to TAZ
lscog_taz_pop <- lscog_taz |>
  sf::st_join(
    lscog_pop_hh_emp |> sf::st_centroid(of_largest_polygon = TRUE)
  ) |>
  dplyr::group_by(
    ID,
    Area,
    TAZ_ID,
    COUNTY,
    AREA_TYPE,
    COUNTYID,
    .drop = FALSE
  ) |>
  dplyr::summarize(
    .groups = "drop",
    # Population and Household Size
    TOTPOP = sum(TOTPOP, na.rm = TRUE),
    GQPOP = sum(GQPOP, na.rm = TRUE),
    HHPOP = sum(HHPOP, na.rm = TRUE),
    HH = sum(HH, na.rm = TRUE),
    HH_1 = sum(HH_1, na.rm = TRUE),
    HH_2 = sum(HH_2, na.rm = TRUE),
    HH_3 = sum(HH_3, na.rm = TRUE),
    HH_4 = sum(HH_4, na.rm = TRUE),
    DU = sum(DU, na.rm = TRUE),
    # Household Income
    INC_14999 = sum(INC_14999, na.rm = TRUE),
    INC_49999 = sum(INC_49999, na.rm = TRUE),
    INC_50000 = sum(INC_50000, na.rm = TRUE),
    # Employment
    TOTAL_EMP = sum(TOTAL_EMP, na.rm = TRUE),
    AGR_FOR_FI = sum(AGR_FOR_FI, na.rm = TRUE),
    MINING = sum(MINING, na.rm = TRUE),
    CONSTRUCTI = sum(CONSTRUCTI, na.rm = TRUE),
    MANUFACTUR = sum(MANUFACTUR, na.rm = TRUE),
    TRANSP_COM = sum(TRANSP_COM, na.rm = TRUE),
    WHOLESALE = sum(WHOLESALE, na.rm = TRUE),
    RETAIL = sum(RETAIL, na.rm = TRUE),
    FIRE = sum(FIRE, na.rm = TRUE),
    SERVICES = sum(SERVICES, na.rm = TRUE),
    PUBLIC_ADM = sum(PUBLIC_ADM, na.rm = TRUE)
  )

lscog_taz_pop
```

#### Python

```{python}

# Aggregate population, households, and employment to TAZ
lscog_taz_pop = (lscog_taz
    .sjoin(
        lscog_pop_hh_emp.assign(geometry=lscog_pop_hh_emp.geometry.centroid).to_crs(lscog_taz.crs),
        how='left'
    )
    .groupby(['ID', 'Area', 'TAZ_ID', 'COUNTY', 'AREA_TYPE', 'COUNTYID'], 
             as_index=False, dropna=False)
    .agg({
        # Population and Household Size
        'TOTPOP': lambda x: x.sum(skipna=True),
        'GQPOP': lambda x: x.sum(skipna=True),
        'HHPOP': lambda x: x.sum(skipna=True),
        'HH': lambda x: x.sum(skipna=True),
        'HH_1': lambda x: x.sum(skipna=True),
        'HH_2': lambda x: x.sum(skipna=True),
        'HH_3': lambda x: x.sum(skipna=True),
        'HH_4': lambda x: x.sum(skipna=True),
        'DU': lambda x: x.sum(skipna=True),
        # Household Income
        'INC_14999': lambda x: x.sum(skipna=True),
        'INC_49999': lambda x: x.sum(skipna=True),
        'INC_50000': lambda x: x.sum(skipna=True),
        # Employment
        'TOTAL_EMP': lambda x: x.sum(skipna=True),
        'AGR_FOR_FI': lambda x: x.sum(skipna=True),
        'MINING': lambda x: x.sum(skipna=True),
        'CONSTRUCTI': lambda x: x.sum(skipna=True),
        'MANUFACTUR': lambda x: x.sum(skipna=True),
        'TRANSP_COM': lambda x: x.sum(skipna=True),
        'WHOLESALE': lambda x: x.sum(skipna=True),
        'RETAIL': lambda x: x.sum(skipna=True),
        'FIRE': lambda x: x.sum(skipna=True),
        'SERVICES': lambda x: x.sum(skipna=True),
        'PUBLIC_ADM': lambda x: x.sum(skipna=True),
        'geometry': 'first'
    })
)

lscog_taz_pop = gpd.GeoDataFrame(lscog_taz_pop, crs=lscog_taz.crs)
lscog_taz_pop
```
:::

## School and college enrollment

The school enrollment combination merges public and private educational institution data into a unified dataset for comprehensive coverage of student populations.

:::panel-tabset

#### R

```{r}
# Combine school enrollment data
lscog_sch_enroll <- dplyr::bind_rows(
  lscog_pub_sch_enroll,
  lscog_pvt_sch_enroll
) |>
  dplyr::mutate(
    STUDENT_COUNT = dplyr::coalesce(STUDENT_COUNT_PUB, 0) +
      dplyr::coalesce(STUDENT_COUNT_PVT, 0),
    TEACHER_COUNT = dplyr::coalesce(TEACHER_COUNT_PUB, 0) +
      dplyr::coalesce(TEACHER_COUNT_PVT, 0)
  )

lscog_sch_enroll
```

#### Python

```{python}

# Combine school enrollment data
lscog_sch_enroll = pd.concat([
    lscog_pub_sch_enroll.assign(
        STUDENT_COUNT=lscog_pub_sch_enroll['STUDENT_COUNT_PUB'],
        TEACHER_COUNT=lscog_pub_sch_enroll['TEACHER_COUNT_PUB']
    ),
    lscog_pvt_sch_enroll.assign(
        STUDENT_COUNT=lscog_pvt_sch_enroll['STUDENT_COUNT_PVT'],
        TEACHER_COUNT=lscog_pvt_sch_enroll['TEACHER_COUNT_PVT']
    )
])

lscog_sch_enroll
```

:::

The subsequent TAZ aggregation counts total student enrollment within each zone, providing essential data for modeling education-related trip patterns and supporting specialized trip generation rates for school-based travel.

:::panel-tabset

#### R

```{r}
# count the number of school enrollment within each TAZ
lscog_taz_enroll <- lscog_taz |>
  sf::st_join(lscog_sch_enroll) |>
  dplyr::group_by(
    ID,
    Area,
    TAZ_ID,
    COUNTY,
    AREA_TYPE,
    COUNTYID,
    .drop = FALSE
  ) |>
  dplyr::summarize(
    .groups = "drop",
    STUDENT_COUNT = sum(STUDENT_COUNT, na.rm = TRUE)
  )

lscog_taz_enroll
```

#### Python

```{python}

# count the number of school enrollment within each TAZ
lscog_taz_enroll = lscog_taz.sjoin(lscog_sch_enroll, how='left') \
    .groupby(['ID', 'Area', 'TAZ_ID', 'COUNTY', 'AREA_TYPE', 'COUNTYID'], as_index=False) \
    .agg({'STUDENT_COUNT': 'sum'})
    
lscog_taz_enroll
```

:::

# Combine into a single data

This step integrates all TAZ-level socioeconomic datasets into a comprehensive base year file for travel demand modeling. This consolidated dataset contains all essential variables organized in a standardized format with geographic identifiers, demographic characteristics, employment data, and educational enrollment totals for each traffic analysis zone in the study region.

:::panel-tabset

#### R

```{r}
# Combine population, household, employments, and school enrollment
lscog_se_base <- lscog_taz_pop |>
  dplyr::left_join(
    lscog_taz_enroll |> sf::st_drop_geometry(),
    by = dplyr::join_by(ID, Area, TAZ_ID, COUNTY, AREA_TYPE, COUNTYID)
  ) |>
  dplyr::select(
    ID,
    Area,
    TAZ_ID,
    COUNTY,
    AREA_TYPE,
    COUNTYID,
    INC_14999,
    INC_49999,
    INC_50000,
    TOTPOP,
    GQPOP,
    HHPOP,
    HH,
    HH_1,
    HH_2,
    HH_3,
    HH_4,
    DU,
    dplyr::everything()
  )

lscog_se_base
```

#### Python

```{python}

# Define the column order
field_order = [
    'ID', 'Area', 'TAZ_ID', 'COUNTY', 'AREA_TYPE', 'COUNTYID',
    'INC_14999', 'INC_49999', 'INC_50000',
    'TOTPOP', 'GQPOP', 'HHPOP',
    'HH', 'HH_1', 'HH_2', 'HH_3', 'HH_4', 'DU'
]

# Combine population, household, employments, and school enrollment
lscog_se_base = lscog_taz_pop.merge(
    lscog_taz_enroll,
    on=['ID', 'Area', 'TAZ_ID', 'COUNTY', 'AREA_TYPE', 'COUNTYID'],
    how='left'
)

lscog_se_base = lscog_se_base[field_order + 
                              list(lscog_se_base.columns.difference(field_order))]

lscog_se_base
```

:::

# Data checks and validation

The validation process ensures data integrity and consistency across all socioeconomic variables through systematic checks of categorical totals. These validation steps identify any discrepancies between aggregate totals and component categories that may have occurred during the interpolation or aggregation processes.

## Household size total

This validation confirms that the total household count matches the sum of all household size categories for each TAZ. Any discrepancies indicate potential issues in the household size distribution that require investigation and correction before model implementation.

:::panel-tabset

#### R
```{r}
# check the sum of household by household size
lscog_se_base |>
  dplyr::filter(HH != (HH_1 + HH_2 + HH_3 + HH_4)) |>
  nrow()
```

#### Python

```{python}

# check the sum of household by household size
lscog_se_base[
    lscog_se_base['HH'] != (lscog_se_base['HH_1'] + lscog_se_base['HH_2'] +
                            lscog_se_base['HH_3'] + lscog_se_base['HH_4'])
].shape[0]
```
:::

## Household income total

The income category validation verifies that household totals equal the sum of all three income brackets across all zones. This check ensures the integrity of the income distribution data following the proportional allocation methodology applied during the ACS interpolation process.

:::panel-tabset

#### R

```{r}
# check the sum of household by income level
lscog_se_base |>
  dplyr::filter(HH != (INC_14999 + INC_49999 + INC_50000)) |>
  nrow()
```

#### Python

```{python}

# check the sum of household by income level
lscog_se_base[
    lscog_se_base['HH'] != (lscog_se_base['INC_14999'] + lscog_se_base['INC_49999'] +
                            lscog_se_base['INC_50000'])
].shape[0]
```

:::

## Employment categories

The employment validation confirms that total employment equals the sum of all industry sector categories for each TAZ. This comprehensive check validates the LEHD data integration and ensures that no employment is lost or double-counted during the sectoral disaggregation process.RetryClaude can make mistakes. Please double-check responses.

:::panel-tabset

#### R

```{r}
# check the sum of employment by categories
lscog_se_base |>
  dplyr::filter(
    TOTAL_EMP !=
      (AGR_FOR_FI +
        MINING +
        CONSTRUCTI +
        MANUFACTUR +
        TRANSP_COM +
        WHOLESALE +
        RETAIL +
        FIRE +
        SERVICES +
        PUBLIC_ADM)
  ) |>
  nrow()
```

#### Python

```{python}

# check the sum of employment by categories
lscog_se_base[
    lscog_se_base['TOTAL_EMP'] != (
        lscog_se_base['AGR_FOR_FI'] +
        lscog_se_base['MINING'] +
        lscog_se_base['CONSTRUCTI'] +
        lscog_se_base['MANUFACTUR'] +
        lscog_se_base['TRANSP_COM'] +
        lscog_se_base['WHOLESALE'] +
        lscog_se_base['RETAIL'] +
        lscog_se_base['FIRE'] +
        lscog_se_base['SERVICES'] +
        lscog_se_base['PUBLIC_ADM']
    )
].shape[0]
```
:::

# Export the final data

The final export creates the complete TAZ-level socioeconomic dataset in both tabular and spatial formats for direct integration into the travel demand model. This comprehensive dataset serves as the primary input for trip generation, providing all necessary demographic, economic, and educational variables organized by traffic analysis zone for the LSCOG regional modeling system.

**Export as CSV flat file**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as CSV
lscog_se_base |>
  sf::st_drop_geometry() |>
  readr::write_csv(
    file.path(root, "Task 1 TDM Development/Base Year/_raw/lscog_se_taz.csv"),
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as CSV
lscog_se_base |>
  gpd.GeoDataFrame.to_csv(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_raw" / "lscog_se_taz.csv",
    index=False
  )
```

:::

**Export as Geodatabase layer**

:::panel-tabset

#### R

```{r}
#| eval: false

# Export as GDB
lscog_se_base |>
  sf::write_sf(
    file.path(
      root,
      "Task 1 TDM Development/Base Year/_gis/LSCOG_2020Base_SE.gdb"
    ),
    layer = "lscog_se_base",
    append = FALSE
  )
```

#### Python

```{python}
#| eval: false

# Export as GDB
lscog_se_base |>
  gpd.GeoDataFrame.to_file(
    Path(root) / "Task 1 TDM Development" / "Base Year" / "_gis" / "LSCOG_2020Base_SE.gdb"
    layer='lscog_se_base',
    driver='FileGDB'
  )
```

:::
